---
title: "How to grow a tree?"
subtitle: "Decision trees"
author: "Michał Burdukiewicz"
date: "03-06-2022"
output: 
  ioslides_presentation:
    theme: "simplex"
---

<link href="https://fonts.googleapis.com/css?family=Montserrat&display=swap" rel="stylesheet">

<style>
slides > slide {
  font-family: 'Montserrat', sans-serif;
}

.center {
  display: block;
  margin-left: auto;
  margin-right: auto;

}


</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Modelling


A modelling algorithm finds a relationship between response variable and input variable.

Questions: can we use number of active days on wikipedia and number of pages created to guess the gender of Wikipedia registered users?

Source: Minguillón, J., Meneses, J., Aibar, E., Ferran-Ferrer, N., and Fàbregues, S. (2021). Exploring the gender gap in the Spanish Wikipedia: Differences in engagement and editing practices. PLOS ONE 16, e0246702. https://doi.org/10.1371/journal.pone.0246702.

## Available data

C_api: gender (male, female, unknown)

pagesWomen: number of edits in pages related to women

NEds: total number of editions

## Basic terms

$y$: response variables.

$X$: input variable with $n$ features.

$$y \sim X$$

$$ y = \beta_1 x_1 + \dots + \beta_i x_i + \dots + \beta_n x_n $$


## Available data

```{r,echo=FALSE,message=FALSE}
library(ggplot2)
library(dplyr)
library(parttree)
library(rpart.plot)
theme_set(theme_bw())


dat <- read.csv("./data.csv") %>% 
  mutate(C_api = factor(C_api))

set.seed(15390)

sampled_dat <- rbind(filter(dat, C_api == "female"),
                     filter(dat, C_api == "male") %>% 
                       slice(sample(x = which(dat[["C_api"]] == "male"), 
                                    size = 2*sum(dat[["C_api"]] == "female")))) %>% 
  mutate(C_api = factor(C_api))

select(sampled_dat, C_api, NEds, pagesWomen) %>% 
  head() %>% 
  knitr::kable()

```


## Available data

```{r,echo=FALSE,message=FALSE}

ggplot(sampled_dat, aes(x = NEds, y = pagesWomen, color = C_api)) +
  geom_point() + 
  facet_wrap(~ C_api)

```

## Without faceting

```{r,echo=FALSE,message=FALSE}

ggplot(sampled_dat, aes(x = NEds, y = pagesWomen, color = C_api)) +
  geom_point() 

```

## Partitioning

Can we divide data into a subsets based on the value of NEds and pagesWomen to distinguish between male and female?


## Partitioning

```{r,echo=FALSE,message=FALSE}

wiki_tree  <- rpart(C_api ~ NEds + pagesWomen, data = sampled_dat, model = TRUE)

ggplot(sampled_dat, aes(x = NEds, y = pagesWomen, color = C_api)) +
  geom_point() +
  geom_parttree(data = wiki_tree, aes(fill = C_api), alpha = 0.1) +
  theme_minimal()

```

## Partitioning

Cen we represent the partition in a simpler way?

```{r,echo=FALSE,message=FALSE}

rpart.plot(wiki_tree)

```


## How to grow it?

$$ y = \beta_1 x_1 + \dots + \beta_i x_i + \dots + \beta_n x_n $$

$M$: number of regions

## How to grow it?

```{r,echo=FALSE,message=FALSE}

ggplot(sampled_dat, aes(x = NEds, y = pagesWomen, color = C_api)) +
  geom_point() +
  geom_parttree(data = wiki_tree, aes(fill = C_api), alpha = 0.1) +
  theme_minimal()

```

## How to grow it?

$$ y = \beta_1 x_1 + \dots + \beta_i x_i + \dots + \beta_n x_n $$

$M$: number of regions

$$ y = \sum^{M}_{m = 1} c_m \left(x \in R_M \right) $$


$$ \hat c_m = \textrm{ave} \left(y_i | x_i \in R_M \right) $$ 


## How to grow it?

Next steps:

 - greedy algorithm for finding optimal splits
 - pruning (collapsing nodes, where their bring very little to the fit)

## Reading more:

Section 9.2, Elements of statistical learning (https://hastie.su.domains/Papers/ESLII.pdf)
